---
title: "Assignment 5: Data Visualization"
author: "Robert Hill"
date: "Spring 2023"
output: pdf_document
geometry: margin=2.54cm
fig_height: 3
fig_width: 4
editor_options: 
  chunk_output_type: console
---

## OVERVIEW

This exercise accompanies the lessons in Environmental Data Analytics on Data Visualization 

## Directions
1.  Rename this file `<FirstLast>_A05_DataVisualization.Rmd` (replacing `<FirstLast>` with your first and last name).
2. Change "Student Name" on line 3 (above) with your name.
3. Work through the steps, **creating code and output** that fulfill each instruction.
4. Be sure your code is tidy; use line breaks to ensure your code fits in the knitted output.
5. Be sure to **answer the questions** in this assignment document.
6. When you have completed the assignment, **Knit** the text and code into a single PDF file.

---

## Set up your session 

1. Set up your session. Load the tidyverse, lubridate, here & cowplot packages, and verify your home directory. Upload the NTL-LTER processed data files for nutrients and chemistry/physics for Peter and Paul Lakes (use the tidy `NTL-LTER_Lake_Chemistry_Nutrients_PeterPaul_Processed.csv` version) and the processed data file for the Niwot Ridge litter dataset (use the `NEON_NIWO_Litter_mass_trap_Processed.csv` version). 

2. Make sure R is reading dates as date format; if not change the format to date.

```{r initialize project}
#1
library(tidyverse);library(lubridate);library(here);library(cowplot)

getwd()

PeterPaul.chem.nutrients <- read.csv(
  "./Data/Processed/NTL-LTER_Lake_Chemistry_Nutrients_PeterPaul_Processed.csv",
  stringsAsFactors = TRUE)

NEON.Litter <- read.csv(
  "./Data/Processed/NEON_NIWO_Litter_mass_trap_Processed.csv",
  stringsAsFactors = TRUE)

#2 
#check if class is date, it is not. It is factor
class(PeterPaul.chem.nutrients$sampledate)
class(NEON.Litter$collectDate)
#change date data to date class
PeterPaul.chem.nutrients$sampledate <- ymd(PeterPaul.chem.nutrients$sampledate)
NEON.Litter$collectDate <- ymd(NEON.Litter$collectDate)
#confirm dates read as a date class
class(PeterPaul.chem.nutrients$sampledate)
class(NEON.Litter$collectDate)
```
## Define your theme
3. Build a theme and set it as your default theme. Customize the look of at least two of the following: 
 * Plot background
 * Plot title
 * Axis labels
 * Axis ticks/gridlines
 * Legend

```{r build plot theme}
#3
library(ggthemes)
my_theme <- theme_base() + 
  theme(
    plot.background = element_rect(
      color='white',
      fill = 'gray'
    ),
    legend.position = "bottom", legend.box = "horizontal",
    legend.title = element_text(face = "bold"),
    plot.title = element_text(hjust = 0.5)
  )
```

## Create graphs
For numbers 4-7, create ggplot graphs and adjust aesthetics to follow best practices for data visualization. Ensure your theme, color palettes, axes, and additional aesthetics are edited accordingly.

4. [NTL-LTER] Plot total phosphorus (`tp_ug`) by phosphate (`po4`), with separate aesthetics for Peter and Paul lakes. Add a line of best fit and color it black. Adjust your axes to hide extreme values (hint: change the limits using `xlim()` and/or `ylim()`).

```{r plot total P vs PO4}
#4 
Phosphorus_Plot <- PeterPaul.chem.nutrients %>%
  ggplot(
    mapping = aes(
      y=tp_ug,
      x=po4,
    )
  ) +
  geom_point(aes(color=lakename)) +
  xlim(0, 50) +
  my_theme +
  labs(
    title = "Phophorus Content",
    y = "Total Phosphorus",
    x = "Phosphate",
    color = ""
  ) +
  geom_smooth(method = "lm", se=FALSE, col="black")


Phosphorus_Plot

```


5. [NTL-LTER] Make three separate boxplots of (a) temperature, (b) TP, and (c) TN, with month as the x axis and lake as a color aesthetic. Then, create a cowplot that combines the three graphs. Make sure that only one legend is present and that graph axes are aligned. 

Tip: R has a build in variable called `month.abb` that returns a list of months;see <https://r-lang.com/month-abb-in-r-with-example>

```{r Create boxplots}
#5
Temp <- ggplot(PeterPaul.chem.nutrients,aes(
  x=factor(month, level=1:12,labels=month.abb), y=temperature_C, color=lakename)) + 
  geom_boxplot() +
  theme(axis.title.x=element_blank(),legend.position = "none") +
  ylab("Temperature")
TP <- ggplot(PeterPaul.chem.nutrients,aes(
  x=factor(month, level=1:12,labels=month.abb), y=tp_ug, color=lakename)) + 
  geom_boxplot() +
  theme(axis.title.x=element_blank(),legend.position = "none") +
  ylab("Total Phosphorus")
TN <- ggplot(PeterPaul.chem.nutrients,aes(
  x=factor(month, level=1:12,labels=month.abb), y=tn_ug, color=lakename)) + 
  geom_boxplot() +
  theme(axis.title.x=element_blank(),legend.position="bottom",legend.box="horizontal") +
  labs(y="Total Nitrogen",color="")
  
plot_grid(
  Temp,TP,TN,
  align="v",
  nrow=3
)

```

Question: What do you observe about the variables of interest over seasons and between lakes? 

> Answer: Nutrient concentrations for Peter Lake are generally higher than that of
>Paul Lake during every month. For the months for which nutrient concentrations
>are available, the median temperature for Peter Lake was lower that that of
>Paul Lake.

6. [Niwot Ridge] Plot a subset of the litter dataset by displaying only the "Needles" functional group. Plot the dry mass of needle litter by date and separate by NLCD class with a color aesthetic. (no need to adjust the name of each land use)

7. [Niwot Ridge] Now, plot the same plot but with NLCD classes separated into three facets rather than separated by color. 

```{r Plot litter}
#6
ggplot(NEON.Litter,aes(collectDate,dryMass,color=nlcdClass)) +
  geom_point(data=subset(NEON.Litter,functionalGroup=="Needles"))

#7
ggplot(NEON.Litter,aes(collectDate,dryMass,color=nlcdClass)) +
  geom_point(data=subset(NEON.Litter,functionalGroup=="Needles")) +
  facet_grid(.~nlcdClass)

```
Question: Which of these plots (6 vs. 7) do you think is more effective, and why? 

> Answer: I think 7 is more effective. Sure, you get less resolution on the x-axis,
>but given the high density of data points near zero, many data points are hidden
>and you may not be able to assess the density of different nlcd class data points
>if going off number 6. You can also better see the trends for each class in 7.